---
# Document information

title: "LSTAT2170 - Time series"
subtitle: "Final Project"

authors:
  - "Lionel Lamy"
  
noma: 
  - "1294-1700"

# ---
# You can remove settings below
# Defaults values are defined

institute: "Université catholique de Louvain"
faculty: "Louvain School of Statistics"
department: ""

context: "Group 12"
date: \today

# ---
colorlinks: yes

linkcolor: "black"
citecolor: "blue"
urlcolor:  "blue"

links-as-notes: no
# ---
output:
  pdf_document:
    template: template/main.tex
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  echo = F,
  eval = T,	
  	
  warning = F,	
  message = F,
  
  	
  out.width= "80%",
  fig.align = "center",	
  fig.path = "resources/figs/"
  
)	

# To display models in formula
# library(equatiomatic)
# equatiomatic::extract_eq(model, use_coefs = TRUE)
```

```{r import, appendix=T}
gas = read.table("./resources/data/gasflorida.txt", header = F)
gas = ts(gas, start=1991, end=1999 + 7/12, frequency = 12)
```

\section{Introduction}



\begin{rmk}
The file describing the data mentions that the time series ends in August 1998, nevertheless the length of the data being 104, the time series ends in August 1999. This temporality will thus be used throughout the document.
\end{rmk}


U.S. Natural Gas State Data 
Deliveries to Residential Consumers
Natural Gas Deliveries to Commercial Consumers (Excluding Vehicle Fuel) in Florida (MMcf)
Monthly - from jan. 1991 to aug. 1998

Remark:"Mcf" means 1,000 cubic feet of natural gas; "Mmcf" means 1,000 Mcf


\section{Conclusion}


<!------>
\appendix

\section{Appendix}
\subsection{Code}

<!-- Remove this part -->
\bigskip
\begin{note*}[about the code]
  All the chunks with the parameter 'appendix' set to TRUE will be outputed below.  
\end{note*}
<!---------End---------->

```{r appendix_code, eval=FALSE, echo=TRUE, attr.source='.numberLines', ref.label=knitr::all_labels(appendix==T)}
```
